\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}

\hypersetup{
    colorlinks=true,
    linkcolor=blue!70!black,
    citecolor=blue!70!black,
    urlcolor=blue!70!black
}

\title{Spectral Graph Features for Point Cloud Classification:\\An Experimental Study}
\author{}
\date{February 2026}

\begin{document}

\maketitle

\begin{abstract}
This report presents a comprehensive experimental study investigating whether Laplacian
eigenvector features from spectral graph theory can improve 3D point cloud classification
on the ModelNet10 and ModelNet40 benchmarks. We evaluate multiple injection
strategies---attention bias, node feature concatenation, learned gated fusion, and
permutation-invariant Deep Sets---across three architecture families: a custom
distance-modulated transformer, PointTransformerV3 (PTv3), and Deep Sets. A systematic
sweep over four eigenvector canonicalization methods and multiple eigenvector counts is
conducted. We further investigate sign-invariant alternatives: Heat Kernel Signatures
(HKS) and eigenvalue spectrum descriptors (Shape-DNA). Statistical analysis of
eigenvector uncanonicalizability across 12,305 shapes provides theoretical grounding for
the observed results. Head-to-head comparison on ModelNet40 with $k=8$ eigenvectors shows
HKS (76.26\%) outperforms eigenvector+Spielman (71.39\%) by 4.87 percentage points,
confirming the advantage of sign-invariant spectral descriptors. An ablation replacing
$v_k(i)^2$ with canonicalized $v_k(i)$ in the exponential weighting (Weighted Eigenvector
Sums, WES) narrows the gap to $\sim$1.5 pp on both MN10 and MN40, revealing that the
multi-scale exponential decay---which downweights high-frequency, hard-to-canonicalize
eigenvectors---accounts for most of HKS's advantage, with squaring providing the remaining
edge. A learnable spectral weighting experiment, replacing the fixed exponential with an MLP,
confirms that increased expressivity does not improve over the fixed kernel (79.41\% vs.\
80.40\% on MN10). At the state-of-the-art level, PTv3 with gated spectral fusion (92.22\%) matches but
does not exceed the geometric-only Pointcept baseline (92.34\%).
\end{abstract}

\tableofcontents
\newpage

%% ============================================================================
\section{Introduction}
%% ============================================================================

The central research question of this study is whether spectral graph features---specifically,
Laplacian eigenvectors computed from point cloud connectivity graphs---can improve 3D point
cloud classification. Eigenvectors of the graph Laplacian encode global geometric structure:
the lowest-frequency eigenvectors capture large-scale connectivity and symmetry patterns,
providing a coordinate-free description of shape that is complementary to raw Euclidean
coordinates.

However, a fundamental challenge arises from the \emph{sign ambiguity} of eigenvectors. For
any eigenvector $\mathbf{v}$ satisfying $L\mathbf{v} = \lambda\mathbf{v}$, the negation
$-\mathbf{v}$ is an equally valid eigenvector. This means that the sign of each eigenvector
is arbitrary, and two isometric shapes may produce eigenvectors with opposite signs. For
neural networks that rely on consistent input representations, this ambiguity poses a
serious obstacle.

Our approach is to systematically test multiple strategies for incorporating spectral
features into point cloud classifiers:
\begin{itemize}[nosep]
    \item \textbf{Injection methods:} additive attention bias from spectral distances, node
          feature concatenation of eigenvectors to coordinates, learned gated fusion of
          geometric and spectral embeddings, and permutation-invariant Deep Sets.
    \item \textbf{Canonicalization strategies:} four methods for resolving sign ambiguity
          (Spielman GF(2) system, max-absolute-value sign convention, random sign, and no
          canonicalization).
    \item \textbf{Sign-invariant alternatives:} Heat Kernel Signatures (HKS), which use
          squared eigenvector entries $v_k(i)^2$, and eigenvalue spectrum descriptors
          (Shape-DNA), which bypass eigenvectors entirely.
    \item \textbf{Architecture families:} a custom distance-modulated transformer,
          PointTransformerV3 (PTv3) with the Pointcept training recipe, and Deep Sets.
\end{itemize}

%% ============================================================================
\section{Methods}
%% ============================================================================

We evaluate eleven model configurations spanning three architecture families.

\paragraph{1. DistanceTransformerClassifier.}
A custom transformer in which the attention logits are multiplicatively modulated by a sparse
$k$-nearest-neighbor distance matrix. The input consists of raw 3D coordinates ($d=3$). After
multi-head self-attention with distance modulation, representations are aggregated via masked
mean pooling and passed to a classification head. This serves as the \emph{geometric-only
baseline} for the custom transformer family.

\paragraph{2. SpectralDistanceTransformerClassifier.}
This model extends the distance transformer with an additive per-head spectral bias derived
from pairwise eigenspace distances. The attention computation becomes
\[
    A = \frac{QK^\top}{\sqrt{d}} \odot D_{\text{kNN}} + B_{\text{spectral}},
\]
where $D_{\text{kNN}}$ is the $k$-NN distance matrix and $B_{\text{spectral}}$ is a learned
projection of pairwise distances in the Laplacian eigenspace.

\paragraph{3. Spectral Node Features.}
The same DistanceTransformerClassifier architecture is used, but the input dimension is
increased from 3 to $3 + n_{\text{eigs}}$ by concatenating canonicalized eigenvectors to the
xyz coordinates as node-level features.

\paragraph{4. PTv3 Pointcept Baseline.}
PointTransformerV3 with the full Pointcept training recipe: 6-channel input
(xyz + normals), encoder depths $(2, 2, 2, 6, 2)$, encoder channels
$(32, 64, 128, 256, 512)$, patch size 1024, 4 serialization orders, sparse convolution with
serialized attention, a 3-layer classifier head, cross-entropy plus Lovasz-Softmax loss,
OneCycleLR scheduling, data augmentations, and 300 training epochs. This serves as the
\emph{geometric-only baseline} for the PTv3 family.

\paragraph{5. GatedSpectralPTv3Classifier.}
A learned gate is inserted between the PTv3 embedding layer and the encoder. Let
$\mathbf{h}_{\text{geo}}$ denote the sparse-convolution embedding of geometric features and
$\mathbf{h}_{\text{spec}}$ denote a projected spectral embedding (Linear $\to$ LayerNorm
$\to$ GELU applied to eigenvectors). A gating vector is computed as
\[
    \mathbf{g} = \sigma\!\bigl(\mathrm{FC}([\mathbf{h}_{\text{geo}} \| \mathbf{h}_{\text{spec}}])\bigr),
\]
and the fused representation is
$\mathbf{h} = \mathbf{g} \odot \mathbf{h}_{\text{geo}} + (1 - \mathbf{g}) \odot \mathbf{h}_{\text{spec}}$.

\paragraph{6. DeepSetsClassifier (Eigenvectors).}
A permutation-invariant Deep Sets architecture~\cite{zaheer2017deep} operating on per-point
features. Each point's features (optionally xyz coordinates concatenated with canonicalized
eigenvector entries) are passed through a shared $\phi$-network (3-layer MLP). The
eigenvector features are scaled by $1/\sqrt{\lambda_k}$ before input to normalize the
energy of each spectral component. A masked mean pool aggregates point-level
representations, and a $\rho$-network (3-layer MLP) produces class logits.

\paragraph{7. DeepSetsClassifier + Eigenvalue Spectrum.}
The same Deep Sets architecture with an additional global branch: the vector of sorted
eigenvalues $(\lambda_1, \ldots, \lambda_k)$ is processed through a separate MLP and
concatenated to the pooled per-point representation before the $\rho$-network. This
provides a global Shape-DNA descriptor that is entirely free of eigenvector sign ambiguity.

\paragraph{8. HKS Deep Sets.}
A Deep Sets model using Heat Kernel Signature (HKS) features instead of raw eigenvectors.
HKS is defined as:
\[
    \text{HKS}(i, t) = \sum_k \exp(-\lambda_k \cdot t) \cdot v_k(i)^2,
\]
which is inherently sign-invariant since it uses $v_k(i)^2$. HKS features are computed
at $T$ log-spaced time scales, producing a $T$-dimensional per-point descriptor. The
graph Laplacian uses Gaussian kernel weights and symmetric normalization.

\paragraph{9. Exponentially-Weighted Eigenvector Sums (WES).}
To isolate the contribution of squaring in HKS, we define a non-squared variant:
\[
    \text{WES}(i, t) = \sum_k \exp(-\lambda_k \cdot t) \cdot v_k(i),
\]
which uses the same exponential multi-scale weighting as HKS but retains the raw
(canonicalized) eigenvector entries $v_k(i)$ instead of $v_k(i)^2$. This variant is
\emph{not} sign-invariant and requires eigenvector canonicalization. It tests whether the
exponential weighting alone---which downweights high-frequency eigenvectors where sign
ambiguity is worst---can compensate for imperfect canonicalization.

\paragraph{10. Learnable Spectral Weighting (Deep Sets).}
This model replaces the fixed exponential kernel $e^{-\lambda_k t}$ in HKS with a learned MLP
that maps the full eigenvalue spectrum $(\lambda_1, \ldots, \lambda_K)$ to a weighting matrix
$W \in \mathbb{R}^{K \times T}$. The per-point spectral features are then computed as:
\[
    f(i, t) = \sum_{k=1}^{K} W_{k,t} \cdot g_k(i),
\]
where $g_k(i) = v_k(i)^2$ (squared, sign-invariant) or $g_k(i) = v_k(i)$ (non-squared,
requiring canonicalization). Unlike HKS, the weighting MLP can learn inter-eigenvalue
relationships and non-exponential decay profiles. This tests whether increased expressivity
in the spectral weighting function can improve upon the fixed exponential.

\paragraph{11. PCA Orientation Canonicalization.}
Points are rotated into their PCA frame (principal component alignment). The sign of each
principal axis is then canonicalized using one of three methods (max-abs, Spielman, random).
This tests whether orientation alignment via PCA can substitute for spectral features.

%% ============================================================================
\section{Experimental Setup}
%% ============================================================================

\subsection{Datasets}

\begin{itemize}[nosep]
    \item \textbf{ModelNet10 (MN10):} 4,899 shapes across 10 object categories (3,991 train / 908 test).
    \item \textbf{ModelNet40 (MN40):} 12,311 shapes across 40 categories (9,843 train / 2,468 test).
    \item \textbf{Point sampling:} 512 points for the custom transformer experiments; 1,024
          farthest-point-sampled points with normals for PTv3 and Deep Sets experiments.
    \item \textbf{$k$-NN graph:} $k = 12$ neighbors for transformer and PTv3; $k = 30$ for HKS.
\end{itemize}

\subsection{Spectral Feature Computation}

Spectral features are computed as follows:
\begin{enumerate}[nosep]
    \item Build a symmetric $k$-NN graph from the point cloud coordinates.
    \item Extract the largest connected component (to ensure a valid Laplacian).
    \item Compute the graph Laplacian $L$ (combinatorial or symmetric normalized).
    \item Solve for the smallest non-trivial eigenpairs via ARPACK, excluding the constant
          eigenvector corresponding to $\lambda_0 = 0$.
    \item Apply one of four canonicalization methods to resolve sign ambiguity:
          \begin{itemize}[nosep]
              \item \textbf{Spielman:} Solve a GF(2) linear system to find a consistent sign
                    assignment.
              \item \textbf{Max-abs:} Flip each eigenvector so that its entry with the largest
                    absolute value is positive.
              \item \textbf{Random:} Assign a random sign to each eigenvector.
              \item \textbf{None:} Use eigenvectors as returned by the eigensolver.
          \end{itemize}
\end{enumerate}

\subsection{Hyperparameters}

\paragraph{Custom Transformer.}
$d_{\text{model}} = 128$, $n_{\text{head}} = 8$, 4 transformer layers, feedforward
dimension 256, dropout 0.1, AdamW optimizer with learning rate $10^{-3}$ and weight decay
$10^{-4}$, CosineAnnealingLR scheduler, 100 training epochs, batch size 16.

\paragraph{PTv3.}
Grid size 0.01, drop-path rate 0.3, AdamW optimizer with learning rate $10^{-3}$ (block
parameters: $10^{-4}$), weight decay 0.01, OneCycleLR scheduler with
$\texttt{pct\_start} = 0.05$, cross-entropy plus Lovasz-Softmax loss, 300 training epochs,
batch size 32.

\paragraph{Deep Sets.}
Hidden dimension 256, 3-layer MLPs for both $\phi$ and $\rho$ networks with ReLU
activations, dropout 0.1, AdamW optimizer with learning rate $10^{-3}$, weight decay
$10^{-4}$, CosineAnnealingLR scheduler, 100 training epochs, batch size 32. Default
spectral computation uses $k = 12$ neighbors, combinatorial Laplacian, Spielman
canonicalization, and $n_{\text{eigs}} = 8$.

\paragraph{HKS Deep Sets.}
Same architecture as Deep Sets. Spectral computation uses $k = 30$ neighbors, Gaussian
kernel weighted adjacency, symmetric normalized Laplacian, and log-spaced time scales for
HKS. Two configurations are evaluated: a standard variant ($n_{\text{eigs}} = 32$,
$T = 16$, hidden dim 256, 100 epochs) and an optimized variant ($n_{\text{eigs}} = 32$,
$T = 32$, hidden dim 512, 200 epochs).

\paragraph{Learnable Spectral Weighting Deep Sets.}
Same Deep Sets architecture with $k = 30$ neighbors, Gaussian kernel weighted adjacency,
symmetric normalized Laplacian, $n_{\text{eigs}} = 8$, $T = 32$ output channels, hidden
dim $= 512$, MLP hidden dim $= 128$ (for the spectral weighting MLP), 200 epochs, batch
size $= 32$, lr $= 10^{-3}$, weight decay $= 10^{-4}$, CosineAnnealingLR, xyz included.

\subsection{Spectral Pipeline Corrections}

During the course of experimentation, we identified and corrected three bugs in the
spectral feature computation pipeline:

\begin{enumerate}[nosep]
    \item \textbf{Graph symmetrization.} The original symmetrization
          $A = \frac{1}{2}(A + A^\top)$ averages one-directional edges, halving distances
          for edges present in only one direction. We corrected this to
          $A = \max(A, A^\top)$ (element-wise maximum), preserving true distances.

    \item \textbf{Near-trivial eigenvector leakage.} The original implementation discarded
          only the first eigenvector (index 0). With weighted Laplacians, multiple
          near-zero eigenvalues can appear due to near-disconnected components. We
          introduced a threshold parameter ($10^{-6}$) to filter \emph{all} eigenvalues
          below it.

    \item \textbf{Unnormalized Laplacian.} The combinatorial Laplacian $L = D - A$ has
          eigenvalues that scale with node degree, making features sensitive to local
          point density. We added support for the symmetric normalized Laplacian
          $L_{\text{norm}} = I - D^{-1/2} A D^{-1/2}$, whose eigenvalues lie in $[0, 2]$.
\end{enumerate}

All Deep Sets and HKS experiments reported below use the corrected pipeline.

%% ============================================================================
\section{Results}
%% ============================================================================

\subsection{Main Results}

Table~\ref{tab:main} summarizes classification accuracy across all model configurations on
ModelNet10 and ModelNet40.

\begin{table}[ht]
\centering
\caption{Classification accuracy (\%) on ModelNet10 (MN10) and ModelNet40 (MN40). Best
result in each architecture family is shown in \textbf{bold}.}
\label{tab:main}
\begin{tabular}{lcc}
\toprule
\textbf{Method} & \textbf{MN10} & \textbf{MN40} \\
\midrule
\multicolumn{3}{l}{\emph{Custom Distance Transformer}} \\
\quad Distance-only (k-NN attention, xyz) & \textbf{82.27} & \textbf{77.19} \\
\quad + Spectral attention bias & 68.14 & 65.40 \\
\quad + Spectral node features (canon.) & 68.39 & 65.19 \\
\quad Eigenvector-only (no xyz) & ${\sim}$11 & -- \\
\midrule
\multicolumn{3}{l}{\emph{PointTransformerV3}} \\
\quad PTv3 naive (xyz, no aug, 100ep) & 54.41 & -- \\
\quad PTv3 + spectral (xyz + eigvecs) & 45.87 & -- \\
\quad PTv3 Pointcept (normals + aug, 300ep) & -- & \textbf{92.34} \\
\quad PTv3 Gated Spectral ($k_{\text{eig}}{=}8$, Spielman) & -- & 92.22 \\
\midrule
\multicolumn{3}{l}{\emph{Deep Sets}} \\
\quad HKS + xyz (optimized: $k_{\text{eig}}{=}32$, $T{=}32$, 512d, 200ep) & \textbf{80.62} & -- \\
\quad HKS + xyz ($k_{\text{eig}}{=}32$, $T{=}16$) & 79.74 & -- \\
\quad HKS + xyz ($k_{\text{eig}}{=}8$, $T{=}16$, 512d, 200ep) & -- & \textbf{76.26} \\
\quad Learnable squared ($k_{\text{eig}}{=}8$, $T{=}32$, 512d, 200ep) & 79.41 & 75.00 \\
\quad Learnable + maxabs & 77.64 & 73.58 \\
\quad Eigvecs + xyz v2 (maxabs, $k_{\text{eig}}{=}8$) & 77.09 & -- \\
\quad Eigvecs + xyz v2 (spielman, $k_{\text{eig}}{=}8$) & 76.98 & -- \\
\quad Eigvecs + xyz + spectrum ($k_{\text{eig}}{=}8$, spielman) & 75.88 & 71.39 \\
\quad Eigvecs + xyz + spectrum ($k_{\text{eig}}{=}32$, maxabs, weighted) & 72.80 & -- \\
\quad PCA (maxabs) & 64.98 & -- \\
\quad PCA (xyz frame only) & 64.54 & -- \\
\quad Eigvecs + xyz (fixed, maxabs) & 60.35 & -- \\
\quad Eigvecs + xyz (fixed, spielman) & 59.03 & -- \\
\midrule
\multicolumn{3}{l}{\emph{Ablations (no xyz)}} \\
\quad HKS only (no xyz) & 31.28 & -- \\
\quad Eigenvalue spectrum only & 22.69 & -- \\
\quad Eigvecs only & ${\sim}$20 & -- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Head-to-Head: HKS vs.\ Eigenvector + Spielman on ModelNet40}

The key comparison of this study is the head-to-head evaluation of two spectral feature
types on ModelNet40 with $k_{\text{eig}} = 8$ eigenvectors. Table~\ref{tab:head2head}
reports the results with full run configurations.

\begin{table}[ht]
\centering
\caption{Head-to-head comparison on ModelNet40, $k_{\text{eig}} = 8$.}
\label{tab:head2head}
\begin{tabular}{lcc}
\toprule
\textbf{Model} & \textbf{Test Acc (\%)} & $\Delta$ \\
\midrule
DeepSets + HKS & \textbf{76.26} & -- \\
DeepSets + Eigvec + Spielman & 71.39 & $-4.87$ \\
\bottomrule
\end{tabular}
\end{table}

\noindent\textbf{DeepSets + HKS configuration:}
\begin{itemize}[nosep]
    \item $k_{\text{eig}} = 8$, $k$-NN neighbors $= 30$, $T = 16$ time scales
    \item Weighted Gaussian kernel adjacency, normalized Laplacian
    \item $\texttt{include\_xyz} = \text{True}$
    \item Hidden dim $= 512$, epochs $= 200$, batch size $= 32$, lr $= 0.001$
    \item Best test accuracy: \textbf{76.26\%}
\end{itemize}

\noindent\textbf{DeepSets + Eigvec + Spielman configuration:}
\begin{itemize}[nosep]
    \item $k_{\text{eig}} = 8$, Spielman canonicalization, fixed $1/\sqrt{\lambda}$ scaling
    \item $\texttt{include\_xyz} = \text{True}$, $\texttt{use\_spectrum} = \text{True}$ (eigenvalue vector appended)
    \item Hidden dim $= 256$, epochs $= 100$, batch size $= 32$, lr $= 0.001$
    \item Best test accuracy: \textbf{71.39\%}
\end{itemize}

HKS outperforms the eigenvector+Spielman approach by 4.87 percentage points. The
sign-invariance of HKS ($v_k(i)^2$ eliminates sign ambiguity by construction) provides a
more robust spectral feature than post-hoc canonicalization of raw eigenvectors.

\subsection{Deep Sets Canonicalization Comparison (ModelNet10)}

Table~\ref{tab:deepsets_canon} compares the three canonicalization methods for Deep Sets
with eigenvector node features on ModelNet10. All runs use $k_{\text{eig}} = 8$,
hidden\_dim $= 256$, 100 epochs.

\begin{table}[ht]
\centering
\caption{Deep Sets eigenvector canonicalization comparison on ModelNet10 (fixed
$1/\sqrt{\lambda}$ scaling, xyz + eigenvectors).}
\label{tab:deepsets_canon}
\begin{tabular}{lcc}
\toprule
\textbf{Canonicalization} & \textbf{Scaling} & \textbf{Test Acc (\%)} \\
\midrule
Max-abs & fixed & \textbf{60.35} \\
Spielman & fixed & 59.03 \\
Random & fixed & 58.26 \\
Spielman & learnable & 58.92 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Improved Deep Sets (v2) with Eigenvectors (ModelNet10)}

The v2 eigenvector models remove the eigenvalue spectrum branch and use a refined
architecture. Both canonicalization methods achieve comparable results, substantially
above the original Deep Sets with eigenvectors.

\begin{table}[ht]
\centering
\caption{Deep Sets v2 eigenvector models on ModelNet10 ($k_{\text{eig}} = 8$, xyz + eigvecs, no spectrum).}
\label{tab:deepsets_v2}
\begin{tabular}{lc}
\toprule
\textbf{Canonicalization} & \textbf{Test Acc (\%)} \\
\midrule
Max-abs & \textbf{77.09} \\
Spielman & 76.98 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{PCA Orientation Canonicalization (ModelNet10)}

All runs: 1,024 points, hidden\_dim $= 256$, 100 epochs.

\begin{table}[ht]
\centering
\caption{PCA-based orientation canonicalization on ModelNet10.}
\label{tab:pca}
\begin{tabular}{lc}
\toprule
\textbf{Sign Method} & \textbf{Test Acc (\%)} \\
\midrule
Max-abs & \textbf{64.98} \\
None (PCA frame only) & 64.54 \\
Spielman & 60.02 \\
Random & 59.03 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Eigenvector-Only Deep Sets (ModelNet10)}

Removing xyz coordinates and relying solely on eigenvector features produces near-random
accuracy across all configurations. Table~\ref{tab:deepsets_noxyz} shows results for
unweighted and weighted Laplacians.

\begin{table}[ht]
\centering
\caption{Deep Sets without xyz coordinates on ModelNet10. Random chance is 10\%.}
\label{tab:deepsets_noxyz}
\begin{tabular}{llc}
\toprule
\textbf{Laplacian} & \textbf{Canonicalization} & \textbf{Test Acc (\%)} \\
\midrule
Unweighted & random & 20.37 \\
Unweighted & spielman & 20.26 \\
Unweighted & maxabs & 19.82 \\
\midrule
Weighted (v2) & maxabs & 17.51 \\
Weighted (v2) & random & 14.98 \\
Weighted (v2) & spielman & 14.98 \\
\midrule
Weighted (v1) & maxabs & 14.54 \\
Weighted (v1) & random & 14.32 \\
Weighted (v1) & spielman & 14.10 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Sign-Invariant Spectral Features}

Table~\ref{tab:sign_invariant} compares HKS and eigenvalue spectrum features on ModelNet10,
including the optimized HKS variant.

\begin{table}[ht]
\centering
\caption{Sign-invariant spectral features on ModelNet10 (Deep Sets).}
\label{tab:sign_invariant}
\begin{tabular}{lcccc}
\toprule
\textbf{Configuration} & $k_{\text{eig}}$ & $T$ & \textbf{Hidden} & \textbf{Test Acc (\%)} \\
\midrule
HKS + xyz (optimized) & 32 & 32 & 512 & \textbf{80.62} \\
HKS + xyz (standard) & 32 & 16 & 256 & 79.74 \\
Eigvecs + xyz + spectrum & 8 & -- & 256 & 75.88 \\
HKS only (no xyz) & 32 & 16 & 256 & 31.28 \\
Eigenvalue spectrum only & 8 & -- & 256 & 22.69 \\
\midrule
\emph{For reference:} & & & & \\
\quad Distance-only transformer & -- & -- & -- & 82.27 \\
\quad Deep Sets eigvecs + xyz (best) & 8 & -- & 256 & 60.35 \\
\quad Deep Sets v2 eigvecs + xyz (best) & 8 & -- & 256 & 77.09 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Effect of Number of Eigenvectors}

Table~\ref{tab:neigs} compares $k_{\text{eig}} = 8$ vs.\ $k_{\text{eig}} = 32$ for
eigenvector+spectrum models on ModelNet10.

\begin{table}[ht]
\centering
\caption{Effect of number of eigenvectors on Deep Sets + Spectrum (MN10).}
\label{tab:neigs}
\begin{tabular}{lccccc}
\toprule
$k_{\text{eig}}$ & \textbf{Canon.} & \textbf{Weighted} & \textbf{Hidden} & \textbf{Epochs} & \textbf{Test Acc (\%)} \\
\midrule
8 & Spielman & No & 256 & 100 & \textbf{75.88} \\
32 & Max-abs & Yes & 256 & 100 & 72.80 \\
\bottomrule
\end{tabular}
\end{table}

Increasing $k_{\text{eig}}$ from 8 to 32 reduces accuracy by 3.1 percentage points,
consistent with the observation that higher-frequency eigenvectors are harder to
canonicalize (see Section~\ref{sec:uncanon}).

\subsection{Spielman Canonicalization Sweep (Transformer)}

To determine whether the eigenvector sign ambiguity can be resolved through better
canonicalization, we performed a systematic sweep over four canonicalization methods
(Spielman, max-abs, random, none) and multiple eigenvector counts $k \in \{4, 8, 16\}$ on
ModelNet10.

\begin{table}[ht]
\centering
\caption{Spielman canonicalization sweep on ModelNet10 (unweighted Laplacian). Best
validation and test accuracy per model--canonicalization pair.}
\label{tab:sweep_unweighted}
\footnotesize
\begin{tabular}{llccc}
\toprule
\textbf{Model} & \textbf{Canon.} & \textbf{$k$} & \textbf{Val Acc (\%)} & \textbf{Test Acc (\%)} \\
\midrule
Node (Dist.\ Transf.) & maxabs & 16 & 73.56 & \textbf{67.07} \\
Node & spielman & 8 & 74.81 & 66.85 \\
Node & none & 4 & 72.68 & 66.63 \\
Node & random & 16 & 73.68 & 66.52 \\
\midrule
Spectral (Std.\ Transf.) & spielman & 8 & 31.95 & 19.49 \\
Spectral & none & 16 & 30.11 & 18.41 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[ht]
\centering
\caption{Spielman canonicalization sweep on ModelNet10 (weighted Laplacian). Best
validation and test accuracy per model--canonicalization pair.}
\label{tab:sweep_weighted}
\footnotesize
\begin{tabular}{llccc}
\toprule
\textbf{Model} & \textbf{Canon.} & \textbf{$k$} & \textbf{Val Acc (\%)} & \textbf{Test Acc (\%)} \\
\midrule
Node & maxabs & 16 & 76.13 & \textbf{66.35} \\
Node & none & 4 & 76.42 & 65.33 \\
\midrule
Spectral & maxabs & 4 & 34.39 & 19.65 \\
\bottomrule
\end{tabular}
\end{table}

The results are striking in their uniformity: across all four canonicalization strategies,
test accuracy for the node model varies by less than 0.6 percentage points (66.52--67.07\%).
The Spielman method, despite its theoretical motivation, offers no meaningful advantage over
the trivial max-abs heuristic. The spectral attention bias model collapses to near-chance
accuracy (${\sim}$19\%) regardless of canonicalization.

\subsection{Eigenvector Uncanonicalizability Analysis}
\label{sec:uncanon}

To understand \emph{why} canonicalization fails, we analyzed the uncanonicalizability of
eigenvectors across 12,305 shapes from ModelNet40.

\begin{table}[ht]
\centering
\caption{Eigenvector uncanonicalizability statistics across ModelNet40 (12,305 shapes).}
\label{tab:uncanon}
\begin{tabular}{lr}
\toprule
\textbf{Statistic} & \textbf{Value} \\
\midrule
Shapes analyzed & 12,305 \\
Points per shape & 512 \\
Eigenvectors computed & 20 \\
$k$-NN neighbors & 12 \\
Avg.\ uncanonicalizable eigvecs / shape & 10.41 (std 4.46) \\
Avg.\ multiplicity-1 eigenvalues & 18.96 / 20 \\
Threshold & 0.221 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[ht]
\centering
\caption{Per-index eigenvector uncanonicalizability rates (\%) on ModelNet40.}
\label{tab:uncanon_rates}
\begin{tabular}{lccccccc}
\toprule
\textbf{Eigvec index} & 1 & 2 & 3 & 5 & 10 & 15 & 20 \\
\midrule
\textbf{Uncanon.\ rate (\%)} & \textbf{0.7} & 40.4 & 43.4 & 49.2 & 56.8 & 58.0 & \textbf{61.4} \\
\bottomrule
\end{tabular}
\end{table}

On average, 10.41 out of 20 eigenvectors per shape (52\%) are uncanonicalizable, despite
nearly all eigenvalues having multiplicity one (18.96 out of 20). The uncanonicalizability
rate increases monotonically with eigenvector index, from just 0.7\% for the Fiedler vector
to 61.4\% for the 20th eigenvector.

\subsection{PointTransformerV3 Results (ModelNet40)}

\begin{table}[ht]
\centering
\caption{PTv3 experiments on ModelNet40.}
\label{tab:ptv3}
\begin{tabular}{lcc}
\toprule
\textbf{Model} & \textbf{Test Acc (\%)} & \textbf{Voting Acc (\%)} \\
\midrule
PTv3 Pointcept (normals + aug, 300ep) & \textbf{92.34} & -- \\
PTv3 Gated Spectral ($k_{\text{eig}}{=}8$, Spielman) & 92.22 & 91.82 \\
\bottomrule
\end{tabular}
\end{table}

\noindent\textbf{Gated Spectral PTv3 configuration:}
$k_{\text{eig}} = 8$, Spielman canonicalization, $k$-NN $= 12$, grid\_size $= 0.01$,
drop\_path $= 0.3$, batch size $= 32$, lr $= 0.001$, 300 epochs.

\subsection{Exponentially-Weighted Eigenvector Sums vs.\ HKS}

To isolate whether the advantage of HKS comes from \emph{squaring} ($v_k(i)^2$, which
eliminates sign ambiguity) or from the \emph{exponential multi-scale weighting}
($e^{-\lambda_k t}$, which downweights high-frequency eigenvectors where sign ambiguity is
worst), we compare standard HKS against a non-squared variant: Weighted Eigenvector Sums
(WES), defined as $\text{WES}(i,t) = \sum_k e^{-\lambda_k t} \cdot v_k(i)$. Since WES is
sign-sensitive, we sweep over three canonicalization methods (Spielman, max-abs, random).

All runs use the same architecture and hyperparameters: $k$-NN $= 30$, weighted Gaussian
kernel, normalized Laplacian, $k_{\text{eig}} = 8$, $T = 32$ time scales, hidden dim
$= 512$, 200 epochs, batch size $= 32$, lr $= 0.001$, xyz included.

\begin{table}[ht]
\centering
\caption{Squared (HKS) vs.\ non-squared (WES) exponentially-weighted features.
All runs use $k_{\text{eig}} = 8$, identical architecture and training.}
\label{tab:hks_vs_wes}
\begin{tabular}{llcc}
\toprule
\textbf{Feature} & \textbf{Canonicalization} & \textbf{MN10 (\%)} & \textbf{MN40 (\%)} \\
\midrule
HKS ($v_k^2$, squared) & none (not needed) & \textbf{80.40} & \textbf{76.34} \\
\midrule
WES ($v_k$, non-squared) & Spielman & 79.41 & 74.64 \\
WES ($v_k$, non-squared) & Random & 79.30 & 74.84 \\
WES ($v_k$, non-squared) & Max-abs & 78.74 & 74.19 \\
\midrule
\multicolumn{2}{l}{\emph{Gap: HKS $-$ best WES}} & $1.0$ & $1.5$ \\
\bottomrule
\end{tabular}
\end{table}

\noindent Additionally, with $k_{\text{eig}} = 32$ on MN10 (same config otherwise):
\begin{itemize}[nosep]
    \item HKS: \textbf{80.62\%}; WES max-abs: 79.63\%; WES random: 78.96\%.
\end{itemize}

\textbf{Key observations:}
\begin{enumerate}[nosep]
    \item The exponential weighting substantially closes the gap between canonicalized
          eigenvectors and sign-invariant HKS. Raw canonicalized eigenvectors in Deep Sets
          achieve only $\sim$60\% on MN10 (Table~\ref{tab:deepsets_canon}), while WES
          reaches $\sim$79\%---recovering most of HKS's performance.
    \item HKS (squared) consistently outperforms WES by $\sim$1--2 percentage points on
          both MN10 and MN40, confirming that sign-invariance via squaring provides a
          real advantage beyond what exponential weighting alone achieves.
    \item Canonicalization method has minimal impact within WES: on MN10 the spread is
          0.67 pp (78.74--79.41\%); on MN40 the spread is 0.65 pp (74.19--74.84\%).
    \item The exponential decay $e^{-\lambda_k t}$ acts as a soft sign-invariance
          mechanism: it assigns near-zero weight to high-frequency eigenvectors (large
          $\lambda_k$), which are precisely those with the highest uncanonicalizability
          rates (Section~\ref{sec:uncanon}). At small time scales $t$, high-frequency
          components contribute; at large $t$, only the well-canonicalizable low-frequency
          eigenvectors survive.
\end{enumerate}

\subsection{Learnable Spectral Weighting vs.\ Fixed Exponential}

To test whether a learned weighting function can outperform the fixed exponential kernel in
HKS, we replace $e^{-\lambda_k t}$ with an MLP that maps the eigenvalue spectrum to a
$K \times T$ weighting matrix. We evaluate four variants (squared and three
canonicalization methods for non-squared) on both MN10 and MN40.

\begin{table}[ht]
\centering
\caption{Learnable spectral weighting vs.\ fixed exponential (HKS/WES).
All learnable runs use $k_{\text{eig}} = 8$, $T = 32$, hidden dim $= 512$, 200 epochs.}
\label{tab:learnable}
\begin{tabular}{llcc}
\toprule
\textbf{Feature} & \textbf{Canonicalization} & \textbf{MN10 (\%)} & \textbf{MN40 (\%)} \\
\midrule
\multicolumn{4}{l}{\emph{Learnable weighting (MLP)}} \\
\quad Squared ($v_k^2$) & none (not needed) & \textbf{79.41} & \textbf{75.00} \\
\quad Non-squared ($v_k$) & Max-abs & 77.64 & 73.58 \\
\quad Non-squared ($v_k$) & Random & 76.76 & 73.22 \\
\quad Non-squared ($v_k$) & Spielman & 76.21 & 73.18 \\
\midrule
\multicolumn{4}{l}{\emph{Fixed exponential (for comparison, same $k_{\text{eig}}{=}8$, $T{=}32$, 512d, 200ep)}} \\
\quad HKS ($v_k^2$, squared) & none (not needed) & 80.40 & 76.34 \\
\quad WES ($v_k$, non-squared) & Spielman & 79.41 & 74.64 \\
\quad WES ($v_k$, non-squared) & Max-abs & 78.74 & 74.19 \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Key observations:}
\begin{enumerate}[nosep]
    \item The learnable MLP does \emph{not} improve over the fixed exponential. On MN10,
          learnable squared (79.41\%) trails HKS (80.40\%) by 1.0 pp; on MN40, learnable
          squared (75.00\%) trails HKS (76.34\%) by 1.3 pp.
    \item The squared vs.\ non-squared pattern is consistent with HKS vs.\ WES: learnable
          squared outperforms the best learnable non-squared variant by 1.8 pp on MN10 and
          1.4 pp on MN40, confirming that sign-invariance provides a robust advantage
          regardless of the weighting function.
    \item Among non-squared learnable variants, canonicalization method has modest impact:
          the spread is 1.4 pp on MN10 (76.21--77.64\%) and 0.4 pp on MN40
          (73.18--73.58\%).
    \item The fixed exponential's inductive bias---monotonically downweighting high-frequency
          eigenvectors---appears to be well-matched to the structure of the problem. The
          learnable MLP, despite having strictly more capacity, cannot discover a better
          weighting scheme with the available training data.
\end{enumerate}

%% ============================================================================
\section{Discussion}
%% ============================================================================

\paragraph{Raw eigenvector features consistently hurt.}
Across every injection method tested---attention bias, node feature concatenation, naive
PTv3 concatenation, and Deep Sets---raw eigenvector features degrade or fail to improve
classification accuracy. On the custom transformer, spectral features reduce accuracy by
12--27 percentage points (Table~\ref{tab:main}). Even gated fusion on PTv3 cannot improve
over the geometric-only baseline (92.22\% vs 92.34\%).

\paragraph{The sign ambiguity is fundamental.}
Our uncanonicalizability analysis (Tables~\ref{tab:uncanon}--\ref{tab:uncanon_rates}) reveals
that 40--61\% of eigenvectors cannot be reliably assigned a canonical sign. The
canonicalization sweeps across both transformers
(Tables~\ref{tab:sweep_unweighted}--\ref{tab:sweep_weighted}) and Deep Sets
(Table~\ref{tab:deepsets_canon}) confirm that all canonicalization strategies produce
test accuracies within a narrow band.

\paragraph{Sign-invariant descriptors partially recover spectral signal.}
HKS features achieve 80.62\% on ModelNet10 (optimized) and 76.26\% on ModelNet40---within
2--3 points of the distance-only transformer baseline on MN10 and within 1 point on MN40
(Table~\ref{tab:sign_invariant}). The head-to-head comparison on MN40
(Table~\ref{tab:head2head}) confirms a 4.87 pp advantage for HKS over eigenvector+Spielman,
validating that sign-invariance is the key factor.

\paragraph{Exponential weighting is a soft sign-invariance mechanism.}
The WES ablation (Table~\ref{tab:hks_vs_wes}) reveals that the exponential multi-scale
weighting $e^{-\lambda_k t}$ does most of the heavy lifting in HKS: WES achieves
$\sim$79\% on MN10 and $\sim$75\% on MN40, up from $\sim$60\% for raw eigenvectors---recovering
the majority of the gap to HKS. The remaining $\sim$1--2 pp gap is attributable to the
squaring operation itself. The exponential decay naturally suppresses high-frequency
eigenvectors (large $\lambda_k$) at large time scales, and these are precisely the
eigenvectors with the highest uncanonicalizability rates (61\% for the 20th eigenvector
vs.\ 0.7\% for the Fiedler vector). This frequency-dependent downweighting acts as a soft
form of sign invariance, explaining why canonicalization method barely matters within WES
(spread $<$0.7 pp on both datasets).

\paragraph{Learnable weighting cannot beat fixed exponential.}
Replacing the fixed exponential $e^{-\lambda_k t}$ with a learned MLP that maps the full
eigenvalue spectrum to a weighting matrix does not improve results
(Table~\ref{tab:learnable}). Despite having strictly more capacity, the learned weighting
trails HKS by $\sim$1 pp on both MN10 and MN40. This suggests the exponential kernel's
inductive bias---monotonic suppression of high-frequency components---is a near-optimal
prior for this task, and the limited training data ($\sim$4K--10K shapes) is insufficient
to learn an equally good weighting from scratch.

\paragraph{More eigenvectors $\neq$ better.}
Increasing $k_{\text{eig}}$ from 8 to 32 reduces Deep Sets+spectrum accuracy from 75.88\%
to 72.80\% on MN10 (Table~\ref{tab:neigs}). This is explained by the uncanonicalizability
analysis: higher-frequency eigenvectors have higher uncanonicalizability rates, introducing
more noise.

\paragraph{PCA orientation is a weak canonicalization.}
PCA-based orientation alignment achieves 64.98\% on MN10 (Table~\ref{tab:pca}), below both
eigenvector v2 (77.09\%) and HKS (80.62\%). PCA canonicalizes the coordinate frame rather
than the spectral features, and suffers from its own sign ambiguity for shapes with
near-degenerate principal moments.

\paragraph{Spectral features without coordinates are insufficient.}
Without xyz coordinates, all spectral representations collapse: eigenvectors $\sim$20\%,
HKS 31.28\%, eigenvalues 22.69\% (Table~\ref{tab:deepsets_noxyz}). Spectral features
complement but cannot replace geometric coordinates.

\paragraph{Gated fusion recovers the baseline.}
The GatedSpectralPTv3Classifier achieves 92.22\% on MN40, compared to 92.34\% Pointcept
baseline. The gate effectively suppresses spectral features, confirming they add noise
rather than signal at the PTv3 performance level.

%% ============================================================================
\section{Conclusion}
%% ============================================================================

Raw Laplacian eigenvectors do not provide useful features for point cloud classification
on ModelNet benchmarks. The fundamental eigenvector sign ambiguity cannot be resolved by any
of the four canonicalization strategies tested---Spielman, max-abs, random, and none all
yield statistically indistinguishable results.

However, sign-invariant spectral descriptors tell a more nuanced story. Heat Kernel
Signatures achieve \textbf{80.62\%} on ModelNet10 (optimized) and \textbf{76.26\%} on
ModelNet40---competitive with the distance-only transformer baseline. The head-to-head
comparison on ModelNet40 with $k=8$ eigenvectors confirms HKS outperforms
eigenvector+Spielman by 4.87 percentage points, establishing sign-invariance as the key
factor rather than the choice of canonicalization method.

A key mechanistic insight comes from the WES ablation: removing the squaring from HKS
while retaining the exponential multi-scale weighting recovers $\sim$79\% on MN10 and
$\sim$75\% on MN40, compared to $\sim$60\% for raw eigenvectors. This reveals that the
exponential decay $e^{-\lambda_k t}$---which naturally downweights high-frequency,
hard-to-canonicalize eigenvectors---accounts for most of HKS's advantage. The squaring
operation provides the final $\sim$1--2 pp edge by fully eliminating residual sign noise.
Furthermore, replacing the fixed exponential with a learned MLP does not improve results
(learnable squared: 79.41\% MN10, 75.00\% MN40 vs.\ HKS: 80.40\%, 76.34\%), indicating
that the exponential kernel's inductive bias is near-optimal for this task.

At the state-of-the-art level, PTv3 with the Pointcept recipe achieves 92.34\% on MN40
using only geometric features. A learned gating mechanism achieves parity (92.22\%) by
suppressing spectral features.

Future directions include applying HKS within stronger architectures (PTv3, DGCNN),
SignNet-style basis-invariant networks that learn over eigenvector products
$v_k(i) \cdot v_k(j)$, and tasks where global connectivity is more critical than local
geometry (segmentation, correspondence, retrieval).

\begin{thebibliography}{9}
\bibitem{zaheer2017deep}
M.~Zaheer, S.~Kottur, S.~Ravanbakhsh, B.~P{\'o}czos, R.~Salakhutdinov, and
A.~Smola. Deep Sets. In \emph{Advances in Neural Information Processing Systems}, 2017.
\end{thebibliography}

\end{document}
